% -----------------------------------------------
% Template for ISMIR Papers — ISMIR 2026
% 6+n page policy: 6 pages max technical content
% Additional pages for references, AI usage, ethics
% -----------------------------------------------

\documentclass{article}
\usepackage[submission]{ismir}
\usepackage{amsmath,cite,url}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{booktabs}

\input{paper/preamble}

% Title (IEEE-compliant title case)
\title{Metadata Quality Analysis of {DJ} Software Libraries: A {Rekordbox} {XML} Study Using the Unified {DJ} Metadata Schema}

% Note: double-blind submission — author names anonymized automatically
% For camera-ready: remove [submission] option from ismir package
\oneauthor
  {Anonymous Authors}
  {Anonymous Affiliations\\\texttt{anonymous@ismir.net}}

% Author list for CC-BY license notice
\def\authorname{F. Author, S. Author, and T. Author}

\sloppy

\begin{document}

\maketitle

\begin{abstract}
DJ software libraries contain rich metadata---BPM, musical key, genre, ratings, labels---critical for harmonic mixing, playlist curation, and music information retrieval. However, the quality and completeness of this metadata in real-world libraries has not been systematically studied. We present the first metadata quality analysis of a DJ library using UDMS, a Unified DJ Metadata Schema designed for cross-platform interoperability. Analyzing 636 tracks from a Rekordbox~7.2.8 XML export, we measure field coverage across 13 canonical metadata fields and identify systematic gaps: while audio-derived fields (duration, bitrate, sample rate) achieve 100\% coverage and musical key achieves 93.2\%, genre annotation is incomplete for 36.9\% of tracks and editorial metadata (ratings, labels) is sparse (12--14\%). We characterize a 70-genre taxonomy with significant fragmentation and find a strong bias toward minor keys (88.4\%) consistent with electronic dance music conventions. Our analysis reveals that 39.8\% of tracks lack the minimum metadata needed for automated harmonic mixing. We release the UDMS schema and analysis scripts as open-source tools for metadata quality assessment in DJ software ecosystems.
\end{abstract}

\section{Introduction}

Modern music information retrieval systems and DJ-specific AI tools depend on structured metadata: BPM for beat matching, musical key for harmonic mixing, genre for playlist curation, and ratings for track selection~\citep{bernardes2018harmonic,liebman2015djmc}. When metadata is missing or inconsistent, these systems degrade silently---a DJ relying on automated key detection may produce dissonant transitions, and playlist generators may group incompatible tracks.

DJ software ecosystems are fragmented: Rekordbox (Pioneer DJ/AlphaTheta), Serato DJ, and Native Instruments Traktor each maintain proprietary database formats with incompatible field definitions and export capabilities~\citep{pyrekordbox,serato-parser}. While cross-platform transfer studies have not yet been conducted, a prerequisite question remains unaddressed: \emph{how complete and reliable is the metadata within a single platform's library?}

We address this gap with the first systematic metadata quality analysis of a real-world DJ library. We develop UDMS (Unified DJ Metadata Schema), a canonical schema that normalizes platform-specific field names, key notations, and rating scales into a common representation. Using UDMS, we analyze a 636-track Rekordbox~7.2.8 library spanning electronic, hip-hop, and experimental genres.

Our contributions:
\begin{itemize}
  \item UDMS, a canonical schema normalizing Rekordbox, Serato, and Traktor metadata fields into a unified representation (13 fields, open-source Python implementation).
  \item The first empirical metadata quality analysis of a DJ library: field coverage, completeness tiers, genre taxonomy fragmentation, and key distribution.
  \item Quantification of metadata gaps that impact automated DJ workflows: 36.9\% of tracks lack genre annotation, and 39.8\% lack sufficient metadata for harmonic mixing.
  \item Open-source analysis tools for metadata quality assessment in DJ software ecosystems.
\end{itemize}

\section{Related Work}

\textbf{Music metadata interoperability.} The MusicBrainz project~\citep{musicbrainz} and Discogs~\citep{discogs} provide community-standard schemas for general music metadata but do not capture DJ-specific fields such as BPM, musical key notation, or energy/mood ratings. The Music Meta Ontology~\citep{deberardinis2023metontology} proposes a flexible semantic model for music metadata interoperability across domains---directly relevant to the cross-platform DJ metadata problem. Discogs-VI~\citep{araz2024discogsvi} demonstrates the value of editorial metadata for version identification. Hardjono et al.~\citep{hardjono2019} propose an open music metadata layer using distributed ledgers but do not analyze DJ-platform interoperability.

\textbf{DJ-specific MIR.} Bernardes et al.~\citep{bernardes2018harmonic} formalize harmonic mixing rules based on the Camelot wheel. Zehren et al.~\citep{zehren2020cuepoints} address automatic cue point detection for DJ mixing. Kim et al.~\citep{kim2020djmixes} analyze real-world DJ mixes using subsequence alignment. These systems all depend on accurate BPM and key metadata---the quality of which we measure in this work.

\textbf{Music metadata quality.} TTMR++~\citep{ttmr2024} enriches music descriptions using LLMs and metadata for retrieval, demonstrating that structured metadata improves downstream tasks. Bukey et al.~\citep{bukey2026musicmetadata} investigate LLMs for music captioning, showing the potential for automated metadata enrichment. Ribov and Hagedorn~\citep{ribov2022} study metadata quality in digital archives. We extend this framework to the DJ domain.

\textbf{DJ software ecosystems.} Rekordbox XML exports have been reverse-engineered by the community~\citep{pyrekordbox}. Serato's database format has been analyzed via the Serato-Library-Parser project~\citep{serato-parser}. Traktor's NML format remains partially undocumented. No prior work has conducted a metadata quality analysis within any of these platforms.

\textbf{Audio analysis for metadata.} Key detection algorithms~\citep{silva2025keydetection} and beat tracking systems~\citep{heydari2021beatnet,chang2024beast} provide automated BPM and key annotation. Audio fingerprinting~\citep{ke2005audio,kurth2008efficient,jang2009pairwise} enables track identification for metadata recovery. Zero-shot tagging~\citep{choi2019zeroshot} offers genre classification without platform-specific training.

\section{Unified DJ Metadata Schema (UDMS)}

\subsection{Design Principles}

UDMS is designed around three principles: (1)~\emph{least common denominator}---use field representations supported by all three platforms, (2)~\emph{expressive sufficiency}---capture all DJ-relevant metadata fields, and (3)~\emph{canonical normalization}---convert platform-specific values (key notation, rating scales) to a common representation.

\subsection{Schema Definition}

UDMS defines 13 canonical fields for this analysis (Table~\ref{tab:udms-fields}). Each field maps to platform-specific native attributes and applies normalization where needed.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{llc}
    \toprule
    \textbf{Field} & \textbf{Type} & \textbf{Rekordbox XML} \\
    \midrule
    title          & string  & Name \\
    artist         & string  & Artist \\
    album          & string  & Album \\
    genre          & string  & Genre \\
    bpm            & float   & AverageBpm \\
    key            & string  & Tonality (Camelot) \\
    key\_numeric   & int     & computed from key \\
    rating         & int [0--5] & Rating ($\div$51) \\
    label          & string  & Label \\
    duration\_sec  & int     & TotalTime \\
    bitrate        & int     & BitRate \\
    sample\_rate   & int     & SampleRate \\
    file\_path     & string  & Location \\
    \bottomrule
  \end{tabular}
  \caption{UDMS canonical fields with Rekordbox XML source mappings. Rating uses Rekordbox's 0--255 scale, normalized to 0--5 by dividing by 51.}
  \label{tab:udms-fields}
\end{table}

\subsection{Key Normalization}

Rekordbox exports musical key in Camelot notation (e.g., ``7A'', ``4B''). UDMS preserves this as the canonical representation and computes a numeric fallback (0--23: 0--11 for minor keys, 12--23 for major keys) for programmatic comparison. Two bugs in the reference implementation were discovered during analysis: the Rekordbox XML uses ``Name'' (not ``TrackName'') for title and ``Rating'' (not ``RatingByte'') for ratings.

\section{Methods}

\subsection{Dataset}

We analyzed a Rekordbox~7.2.8 (AlphaTheta) XML export containing 636 tracks. The library spans multiple acquisition sources: DJ playlists (42.3\%), local downloads (24.1\%), monthly download batches (18.4\%), SoulSeek (13.1\%), Rekordbox samplers (1.9\%), and Pioneer DJ demos (0.3\%). File formats: AIFF (71.4\%), FLAC (18.1\%), MP3 (8.6\%), WAV (1.9\%).

\subsection{Analysis Pipeline}

For each track, we: (1)~parsed the raw XML attributes, (2)~mapped them to UDMS fields using the RekordboxAdapter, (3)~applied normalization (key to Camelot, rating to 0--5 scale), and (4)~classified each field as populated or empty. We define a field as \emph{populated} if its UDMS value is non-zero and non-empty (excluding ``Unknown'' for key).

\subsection{Completeness Tiers}

We define five metadata completeness tiers:
\begin{itemize}
  \item \textbf{Minimal}: track title present
  \item \textbf{Basic}: title + artist + BPM
  \item \textbf{Standard}: Basic + musical key
  \item \textbf{Full}: Standard + genre
  \item \textbf{Complete}: Full + label + rating
\end{itemize}

Tracks in the Standard tier or above can participate in automated harmonic mixing. Tracks in the Full tier support genre-based playlist generation.

\section{Results}

\subsection{Field Coverage}

Table~\ref{tab:field-coverage} shows UDMS field coverage across the 636-track library. Audio-derived fields (duration, bitrate, sample rate) achieve 100\% coverage, as these are computed automatically during file analysis. Title is universally present (100\%). Artist and album coverage are high (96.9\% and 96.5\%), with gaps primarily in sampler tracks and unnamed artists.

Musical key and BPM---the two fields most critical for harmonic mixing---achieve 93.2\% and 92.3\% coverage respectively. Genre annotation is the most significant gap, with only 63.1\% of tracks annotated. Editorial metadata (ratings, labels) is sparse at 12.4\% and 13.8\%.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lrrc}
    \toprule
    \textbf{Field} & \textbf{Populated} & \textbf{Total} & \textbf{Coverage} \\
    \midrule
    title          & 636 & 636 & 100.0\% \\
    duration\_sec  & 636 & 636 & 100.0\% \\
    bitrate        & 636 & 636 & 100.0\% \\
    sample\_rate   & 636 & 636 & 100.0\% \\
    file\_path     & 636 & 636 & 100.0\% \\
    artist         & 616 & 636 &  96.9\% \\
    album          & 614 & 636 &  96.5\% \\
    key            & 593 & 636 &  93.2\% \\
    bpm            & 587 & 636 &  92.3\% \\
    genre          & 401 & 636 &  63.1\% \\
    label          &  88 & 636 &  13.8\% \\
    rating         &  79 & 636 &  12.4\% \\
    \bottomrule
  \end{tabular}
  \caption{UDMS field coverage in the 636-track Rekordbox library. Fields sorted by coverage.}
  \label{tab:field-coverage}
\end{table}

\subsection{Metadata Completeness Tiers}

Table~\ref{tab:completeness} shows the distribution of tracks across completeness tiers. Only 60.2\% of tracks reach the ``Full'' tier (title + artist + BPM + key + genre), and only 1.6\% have complete metadata including label and rating. Critically, 39.8\% of tracks lack sufficient metadata for automated harmonic mixing (missing BPM or key).

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lr}
    \toprule
    \textbf{Tier} & \textbf{Tracks} \\
    \midrule
    Minimal (title)              & 636 (100.0\%) \\
    Basic (+ artist + BPM)       & 573 ( 90.1\%) \\
    Standard (+ key)             & 552 ( 86.8\%) \\
    Full (+ genre)               & 383 ( 60.2\%) \\
    Complete (+ label + rating)  &  10 (  1.6\%) \\
    \bottomrule
  \end{tabular}
  \caption{Metadata completeness tiers. Each tier includes all fields from previous tiers.}
  \label{tab:completeness}
\end{table}

\subsection{BPM Distribution}

BPM values (n=587, range 73--175, mean 130.7, SD 14.5) cluster heavily in the 120--149 range (82.1\%), consistent with electronic dance music conventions. The 120--129 BPM bucket contains the most tracks (34.1\%), followed by 130--139 (28.1\%) and 140--149 (19.9\%). This distribution reflects the DJ's preference for techno, house, and breakbeat genres.

\subsection{Musical Key Distribution}

Of 593 keyed tracks, 88.4\% are in minor keys (Camelot ``A'') and 11.6\% in major keys (``B''). This strong minor-key bias is characteristic of electronic dance music, where minor keys convey tension and energy suited to dancefloor contexts~\citep{bernardes2018harmonic}. The most common keys are 4A (F~minor, 11.0\%), 8A (A~minor, 10.6\%), and 6A (G~minor, 10.3\%), with the three most common minor keys accounting for 31.9\% of all keyed tracks.

43 tracks (6.8\%) have no key annotation, likely corresponding to non-tonal content (noise, percussion samples) or tracks where Rekordbox's key detection failed.

\subsection{Genre Taxonomy}

401 tracks carry genre annotation across 70 unique genre strings---a high fragmentation rate indicating inconsistent manual annotation. After grouping similar genres, the library spans: Techno (17.0\%), Breaks/Electro (14.2\%), Dub Techno (12.0\%), House (8.5\%), Dub/Dubwise (7.0\%), Psy Tech (5.7\%), Dubstep (5.2\%), UK Garage/2-Step (4.2\%), and Hardgroove (4.2\%).

The 235 untagged tracks (36.9\%) still carry title (100\%), artist (92.8\%), album (91.5\%), key (87.7\%), and BPM (85.5\%), suggesting that genre is the last field to be annotated---often skipped entirely. This is a significant gap for genre-based playlist generation and automated track recommendation.

\subsection{Impact on Automated DJ Workflows}

We assess metadata readiness for three common DJ workflows:
\begin{itemize}
  \item \textbf{Harmonic mixing} (requires BPM + key): 84.3\% of tracks (536/636) are ready.
  \item \textbf{Genre-based playlists} (requires genre): 63.1\% of tracks are ready.
  \item \textbf{Energy-based sequencing} (requires ratings or energy): 12.4\% of tracks are ready.
\end{itemize}

These gaps mean that automated DJ tools operating on real-world libraries will encounter missing metadata in 16--88\% of cases depending on the workflow.

\section{Discussion}

\subsection{Genre as the Critical Bottleneck}

Genre annotation is the most impactful metadata gap in DJ libraries. At 63.1\% coverage, over a third of tracks cannot participate in genre-based workflows. Moreover, the 70 unique genre strings present a taxonomy problem---manual annotation is inconsistent, with tracks from the same artist or release tagged differently. Automated genre classification using audio features~\citep{choi2019zeroshot,bukey2026musicmetadata} could recover missing annotations and standardize the taxonomy.

\subsection{Minor Key Bias}

The 88.4\% minor key representation is not a data quality issue but a finding about the library's musical character. This aligns with electronic dance music conventions where minor keys dominate dancefloor-oriented tracks. However, the 11.6\% major key tracks are important for set dynamics---DJs use major key tracks for ``release'' moments. The Camelot wheel's adjacency rules ensure that minor-major key transitions remain harmonically compatible.

\subsection{Rating Sparsity}

Only 12.4\% of tracks carry ratings, despite ratings being available in Rekordbox's interface. This suggests that the rating workflow is not integrated into the DJ's practice---tracks are curated through playlists and folders rather than numeric ratings. The UDMS rating normalization (Rekordbox 0--255 to 0--5) produces a clean mapping but is rarely used.

\subsection{Limitations}

This study analyzes a single library from one DJ. Library composition reflects the DJ's personal taste (electronic/bass music focus), limiting generalizability. The 636-track dataset, while substantial, is smaller than the 2,147-track datasets used in commercial DJ software benchmarks. We analyze only Rekordbox XML exports; Serato and Traktor libraries may exhibit different metadata patterns. The UDMS schema does not capture playlist structure, cue points, or loop markers---fields that are important for DJ performance but not stored in the XML export format.

\section{Conclusion}

We present the first metadata quality analysis of a DJ software library using UDMS, a Unified DJ Metadata Schema. Our analysis of 636 tracks from a Rekordbox~7.2.8 export reveals that while audio-derived fields and musical key achieve high coverage (93--100\%), genre annotation---the field most critical for automated playlist generation---is incomplete for 36.9\% of tracks. Editorial metadata (ratings, labels) is sparse across the library.

These findings have practical implications for DJ software developers and MIR researchers: automated tools must gracefully handle missing metadata, and genre enrichment from external sources (MusicBrainz, Discogs, audio classification) is a high-value target for future work. The UDMS schema and analysis tools are released as open-source resources at \url{https://github.com/interfluve-wav/dj-metadata-paper}.

\section*{Acknowledgments}

[Anonymized for review.]

\bibliography{references}

\section{AI Usage Statement}
AI tools were used in the research and writing process for code generation, literature search, and text editing assistance.

\section{Ethics Statement}
This study analyzes a single DJ's personal music library. No human subjects were involved; all metadata was self-generated through normal software use. The dataset contains no personally identifiable information beyond file paths, which are anonymized in the released data.

\end{document}
